{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-10-10 sftrends_manu.Rnw
results minor revision, ready to bulk up discussion

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  Rather than fitting a global model to the entire time series, one regression is fit at every point in the time series.  Observations within a window used to fit each regression are weighted relative to annual, seasonal, and flow distances from the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to each point in the time series. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.

@@

Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium at P8 and C3 showed a consistent increase over time, followed by a reduction beggining in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at P8 and C10 showed increases followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did show any noticeable trends.  Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for C3.  Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun stations, although the reduction in the late 1990s was less pronounced.  By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

@@

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}. The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}. We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}
Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.

Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).  These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}. Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.  Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.  However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.

@@

% obs v flow norm trends
Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.

@@

% Suisun conclusions
As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

>>>>

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

@@

Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.  Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.  Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.  By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).  Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

@@

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply \citep{Jassby08,Schoellhamer11,Cloern12b}, although decreases in phytoplankton by clam biofiltration may have also increased clarity \citep{MacNally10}. The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}. We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}
Although the relationship between phytoplankton and clams have been well described in \ac{sfe} \citep{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion and the relationship of phytoplankton with clam abundance was mediated by flow and climatic variation in recent years.

Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}d) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).   Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (i.e., Figures \ref{fig:clmchl}e and \ref{fig:clmchl}h compared to \ref{fig:clmchl}d and \ref{fig:clmchl}g), such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.  However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.

@@

% overall decreasing trends in nitrogen concentrations, but more subtle variation depending on sites - and by season
%Peripheral high, but opposing trends at some stations - P8 highest for both ammonium and nitrite/nitrage but C3 also high in ammonium but not nitrite/nitrate, C10 low in ammonium but high in nitrite/nitrate

% Suisun increases in ammonium

% obs v flow norm trends
Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.
% what about seasonal changes in the flow-normalized results? Does this demonstrate effects of normal seasonal rpecip on trend intepretations?

@@

% Suisun conclusions
These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.   Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.

In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay.

As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

### 2017-10-10 sftrends_manu.tex
results minor revision, ready to bulk up discussion

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Mon Oct 9 10:41:34 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Mon Oct 9 10:48:31 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).  Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).
